---
title: "Calculation of basic and effective reproduction numbers"
author: "Andrew Tredennick"
date: 2019-02-19
output: html_notebook
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(pomp)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, cache = FALSE)
source("../../code/make-pomp-simulator-function.R")
```

# Motivation
I want to document, once and for all, how we are calculating the effective reproduction number ($R_E$) for the model fitting `pomp` model and the model simulating `pomp` model.
Both models are *SEIR*, but the simulating model includes demography to maintain a constant equilibrium population size, whereas the fitting model includes known population size and more limitted demography (i.e., no deaths).

### Fitting model
The fitting model does not explicitly model the recovered class because we have known population size.
Therefore, we only model unvaccinated births into the susceptible class and do not model deaths from any class.
We calculate $R_E$ at time *t* as:

$$
R_E(t) = \frac{\beta_t}{\gamma} \times \frac{S_t}{N_t},
$$
where $\beta_t$ is tranmission rate at time *t*, $1/\gamma$ is the constant recovery period, $S_t$ is the number of susceptibles at time *t*, and $N_t$ is the total population size (known) at time *t*.

```{r re-fitting}
get_re <- function(beta, gamma = (1/5), s_over_n){
  (beta/gamma) * (s_over_n)
}

beta <- seq(100, 700, by = 100)
s_over_n <- seq(0, 0.001, length.out = 20)

res <- tibble() 
for(do_beta in beta){
  for(do_frac in s_over_n){
    tmp <- get_re(beta = do_beta, s_over_n = do_frac)
    res %>%
      bind_rows(tibble(
        beta = do_beta,
        s_over_n = do_frac,
        re = tmp
      )) -> res
  }
}

ggplot(res, aes(x = s_over_n, y = re, color = as.factor(beta))) +
  geom_line() +
  labs(x = expression(italic(S/N)), y = expression(italic(R)[E])) +
  ggthemes::scale_color_colorblind(name = expression(beta)) +
  ggtitle("Effective reproduction number for fitting model")
```

### Simulating model
The simulating model includes vaccinated births in the recovered class and deaths from all *SEIR* compartments.
We set the birth rate ($\mu$) to equal the death rate ($\nu$; $\mu = \nu = 0.05$) to maintain a relatively constant population size (births and deaths are still stochastic, which is why population size will still vary).
Following an online doc ([here](https://web.stanford.edu/~jhj1/teachingdocs/Jones-on-R0.pdf)), we calculate $R_0$ as:

$$
R_0 = \frac{\eta \beta}{ (\eta + \nu)(\gamma + \nu)} 
$$

and $R_E(t)$ as:

$$
R_E(t) = \frac{\eta \beta_t }{ (\eta + \nu)(\gamma + \nu)} \times \frac{S_t}{N_t}
$$
where parameters are as in the previous $R_E$ equation with the addition of the latent period ($1/\eta$), birth rate ($\mu$), and death rate ($\nu$).

```{r re-simulating}
get_re_sim <- function(beta, eta = 1/8, mu = 0.05, nu = 0.05,
                       gamma = 1/5, s_over_n){
  ((eta*beta) / (eta+nu)*(gamma+nu)) * s_over_n
}

beta <- seq(100, 700, by = 100)
s_over_n <- seq(0, 0.001, length.out = 20)

res <- tibble() 
for(do_beta in beta){
  for(do_frac in s_over_n){
    tmp <- get_re_sim(beta = do_beta, s_over_n = do_frac)
    res %>%
      bind_rows(tibble(
        beta = do_beta,
        s_over_n = do_frac,
        re = tmp
      )) -> res
  }
}

ggplot(res, aes(x = s_over_n, y = re, color = as.factor(beta))) +
  geom_line() +
  labs(x = expression(italic(S/N)), y = expression(italic(R)[E])) +
  ggthemes::scale_color_colorblind(name = expression(beta)) +
  ggtitle("Effective reproduction number for simulating model")

```